Formal Modelling of a Microcontroller Instruction Set in B

### 2009-06-26 papers/SBMF2009/source/paper.tex

\documentclass{llncs}

@@

\title{A Formal Model of Microcontroller Instruction Set in B}

@@

\begin{abstract}
This paper describes an approach to model the microcontroller platforms. More specifically, it shows details about
the Z80 model. The model has been developed using the B method; which applies math and logic concepts to describe
characteristics from platforms. Therefore, this modelling can be used in platform projects to document, build
simulators, verify properties about the model and verify at assembly level software. The verification
at assembly is the feature more important, because this allows to develop faithful software to the algorithm
language. Thus, the platform model was built for this purpose. Finally, this paper shows some relevant techniques
used in models and to describe some details of our approach.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

@@

This work describes briefly the modelling of microcontroller Z80~\cite{Z80_manual} and some techniques used to build
it. This modelling\footnote{The interested reader in more details is invited to visit our repository
at: http://code.google.com/p/b2asm} use the B method~\cite{Abrial}, because it has a great maturity and advanced
support. Besides, the modelling of microcontroller can extend the level of verification of B method, as it will be
explained.

The B method supports the construction of safety systems model by verification of proofs that guarantees its
correctness. So, an initial abstract model of the system is defined and refined until the implementation model. The
B tools  still have the code generator to programming language, but this feature is not verified. In~\cite{Dantas_SBMF08} is presented an approach to extend the verification of B method until the assembler language.
In this approach, the construction of systems begins from the abstract level and it is formally refined until the
target assembly platform.

Therefore, the first step of this approach is to build the platform modelling, of the microcontroller, in B. Using the
responsibility division mechanism provided by B, basic modules was developed to be used as elementary library to
help the constructing of microcontroller model. The elementary library has many definitions about common concepts used
in the microcontrollers. This library is used by Z80 model and other two model microcontrollers that
are under construction.

The microcontroller modelling can be used in platform projects documentation, simulators construction, consistency
verification and software verification at assembly level. The verification at assembly level is more interesting
utility because it allows the development of trustworthy software by construction, that is extremely desirable in
embedded critical software.

In this paper we introduced the modelling of Z80 microcontroller, since elementary library until assembly
instructions. The paper is structured as follows. Section~\ref{sec:B_method} provides a general introduction to the
B method. Section~\ref{sec:models} presents the elementary libraries and of microcontrollers basic concepts.
Section~\ref{sec:z80} presents the Z80 B modelling. Section ~\ref{sec:Proofs} explain a little about the proofs of
models. The next section (\ref{sec:relatedworks}) discusses the related works. Finally, the last section has the
conclusions.

@@

 Essentially, a B module contains two main parts: a header and the available operations, in the
 figure~\ref{fig:maqB} has a intuitive example. The clause $\mathit{MACHINE}$ has the name of module.  The following two are used to access an external module and create an instance of an external module. The
 $\mathit{VARIABLES}$  declares the variable names. The following declares its types and the restrictions. The
 $\mathit{INITIALIZATION} $  defines the its initial value. The last define the operations that remember procedures
 from programming language.

\begin{figure}
%\begin{small}
$$
\begin{array}{l}
\MACHINE \quad  \mathit{micro} \quad \quad \quad \quad \quad  \quad \quad \quad \quad \INITIALISATION  \mathit{pc}
:= 0\\
\SEES \quad \mathit{TYPES}, \mathit{ALU} \quad \quad \quad \quad \quad \quad \quad \quad \OPERATIONS\\
\INCLUDES    \quad \mathit{MEMORY} \quad \quad \quad \quad \quad \quad \quad \mathit{JMP} \mathit{( jump )} =
\\
\VARIABLES \quad    \mathit{pc}  \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \PRE \mathit{jump}
\in\mathit{INSTRUCTION}\\ 
 \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION}  \quad \ \THEN \mathit{pc} :=
\mathit{jump}\\  
\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
\END \quad  \ldots \\
% \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
%\quad  \ldots
\end{array}
$$
% \begin{array}{l}
% \MACHINE \quad \quad \quad \quad \quad \quad  \quad \quad \quad \quad \INITIALISATION \\
% \quad   \mathit{micro} \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \mathit{pc} := 0\\
% \SEES \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \OPERATIONS\\
% \quad \mathit{TYPES}, \mathit{ALU}\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad \\
% \INCLUDES                \quad \quad \quad \quad \quad \quad \quad \quad \quad   \quad   \mathit{JMP} \mathit{( jump )} = \\
% \quad   \mathit{MEMORY}   \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \PRE\\
% \VARIABLES      \quad \quad \quad \quad \quad \quad \quad \quad \quad  \quad \quad  \mathit{jump} \in
% \mathit{INSTRUCTION}\\ \quad   \mathit{pc}    \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad\quad \quad \THEN\\
% \INVARIANT   \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad   \mathit{pc} := \mathit{jump}\\
% \quad \mathit{pc} \in \mathit{INSTRUCTION} \ \ \quad \quad \quad \quad \quad \quad \END\\
% \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad
% \quad \quad  \ldots
% \end{array}

%\end{small}
\caption{A part of B simplified module}
\label{fig:maqB}
\end{figure}

\section{Overview of Plataform's B Models}
\label{sec:models}

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped in two projects of libraries and other project to model the
platform. Thus, our workspace is arranged through three projects: hardware library, types library and
the project of specific platform, in this case the Z80.

The dependency graph between these projects is depicted in figure~\ref{fig:hardware-definition-graph}
and the next sections will explain each one of them.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency graph between the projects that compose the platform model of Z80.}
\label{fig:hardware-definition-graph}
\end{figure}

@@

First, bits are modelled as the set of integers: $\mathit{BIT} = \mathit{0..1}$. The negation is an unary
function on bits and it is defined as:

@@

Sequences are pre-defined in B, as functions whose the domain is an integer range with lower bound 1 (one). Indices
in bit vectors usually range from 0 (zero) upwards and the model we propose obeys this convention by making an
one-position shift where necessary. This shift is important to use the predefined functions of sequences. We
thus define bit vectors as non-empty sequences of bits, and $\mathit{BIT\_VECTOR}$ is the set of all such sequences:
$
\begin{array}{l}
\mathit{BIT\_VECTOR} = \seq (\mathit{BIT})
\end{array}
$

@@

The \textit{Z80} is an important CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports
158 different instructions. These instructions are classified into these categories: load and exchange; block transfer
and search; arithmetic and logical; rotate and shift; bit manipulation (set, reset, test); jump, call and return; input/output; and basic cpu control.

@@

The Z80 has many register of diferrents types and many instructions. It includes alternative set of accumulator,
flag and general registers. The CPU contains a stack pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two
index registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register ($\mathit{i\_}$), a refresh register
($\mathit{r\_}$), two bits ($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions, a pair of bits to
define the interruption mode ($\mathit{im}$) and the input and output ports ($\mathit{i\_o\_ports}$). Below, its
definitions are specified in the \textit{INVARIANT}.

@@

\section{Proofs}
\label{sec:Proofs}
% [intro/scope] [O que as provas garantem? citar que dificeis, que é necessário buscar técnicas eficientes
% para realizar as provas ]

@@

% [ I need review the english]

@@

 \section{Conclusions}
\label{sec:conclusions} 
This work has shown an approach to the formal modelling of microcontrollers using the B method. This approach
provide interesting results and had a great automatization in the prove process. For instance, it is interesting to
say that: during the building process of the Z80 model we found some errors and ambiguities in the official
manual\cite{Z80_manual}. The platform model can replace (or improve) the documentation used by assembler
programmers. Because the formal model represents the instructions effects and the B method has an easy notation,
similar to pascal language.  Besides, the formal model restricts the definitions to correct typing, uses expressions
well-defined and allows verifying properties of the model.

@@

% [Trabalhos futuros ]
Future works comprise the development of a real study case in the oil area and the project of a compiler to aid the
approach.
%  The next steps of the work comprises the development of a real study case in the oil area and the project of a
% compiler to aid the automatization of verification at the assembly level

@@

 \paragraph{Acknowledges:}
This work received supports from ANP (National Agency of Oil) and CNPq.
% Besides, the researcher Bruno Gomes help in the revision.

>>>>

\documentclass[a4paper]{llncs}

@@

\title{Formal Modelling of a Microcontroller Instruction Set in B}

@@

\begin{abstract}
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform. This work is a contribution towards the extension of the
  B method to handle developments up to assembly level code.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

@@

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for several programming languages, but the result of this translation
cannot be compared by formal means. \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language language. One key component of this approach
is to build, within the framework of the B method, formal modelos of
the instruction set of such assembly languages.

This work gives an overview of the formal modelling of the instruction
set of the Z80 microcontroller~\cite{Z80_manual}\footnote{The
  interested reader in more details is invited to visit our repository
  at: http://code.google.com/p/b2asm.}. Using the responsibility
division mechanism provided by B, auxiliary libraries of basic modules
were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers; besides the Z80 model, it is used by two other
microcontrollers models that are under way.

Other possible uses of a formal model of a micro-controller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

This paper is focused on the presentation of the Z80 model, including
elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short
introduction to the B method. Section~\ref{sec:models} presents the
elementary libraries and the modelling of some elements common to
microcontrollers.  Section~\ref{sec:z80} presents the B model of the
Z80 instruction set. Section ~\ref{sec:Proofs} provides some
information on the proof effort needed to analyze the presented
models. Related work is discussed in
Section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

@@

Essentially, a B module contains two main parts: a header and the
available operations. Figure~\ref{fig:maqB} has a very basic
example. The clause $\mathit{MACHINE}$ has the name of module.  The
following two clauses respectively reference external modules and
create an instance of an external module. The $\mathit{VARIABLES}$
declares the name of the variables that compose the state of the
machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $
sepecifies the initial states. Finally, operations correspond to
the transitions between states of the machine.

\begin{figure}
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{1cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

\section{Model structure and basic components}
\label{sec:models}

We have been developed a reusable set of basic definitions to model
hardware concepts and data types concepts. These definitions are
grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of
the Z80 model. Thus, the workspace is composed of: a hardware library,
a types library and a project for the specific platform, in this case
the Z80. The corresponding dependency diagram is depicted in
Figure~\ref{fig:hardware-definition-graph}; information specific to
each project is presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.jpg}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

@@

First, bits are modelled as a set of integers: $\mathit{BIT} =
\mathit{0..1}$. The negation is an unary function on bits and it is
defined as:

@@

Sequences are pre-defined in B, as functions whose the domain is an
integer range with lower bound 1 (one). Indices in bit vectors usually
range from 0 (zero) upwards and the model we propose obeys this
convention by making an one-position shift where necessary. This shift
is important to use the predefined functions of sequences. We thus
define bit vectors as non-empty sequences of bits, and
$\mathit{BIT\_VECTOR}$ is the set of all such sequences: 
$\mathit{BIT\_VECTOR} = \seq (\mathit{BIT})$.

@@

The \textit{Z80} is a CISC microcontroller developed by
\textit{Zilog}~\cite{Z80_manual}. It supports 158 different
instructions. These instructions are classified into these categories:
load and exchange; block transfer and search; arithmetic and logical;
rotate and shift; bit manipulation (set, reset, test); jump, call and
return; input/output; and basic cpu control.

@@

The Z80 has many register of differents types and many instructions. It includes alternative set of accumulator,
flag and general registers. The CPU contains a stack pointer ($\mathit{sp}$), program counter ($\mathit{pc}$), two
index registers ($\mathit{ix}$ and $\mathit{iy}$), an interrupt register ($\mathit{i\_}$), a refresh register
($\mathit{r\_}$), two bits ($\mathit{iff1}$, $\mathit{iff2}$) used to control the interruptions, a pair of bits to
define the interruption mode ($\mathit{im}$) and the input and output ports ($\mathit{i\_o\_ports}$). Below, its
definitions are specified in the \textit{INVARIANT}.

@@

\section{Proofs}
\label{sec:Proofs}
% [O que as provas garantem? citar que dificeis, que é necessário buscar técnicas eficientes
% para realizar as provas ]

@@

@@

 \section{Conclusions}
\label{sec:conclusions}

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, several ambiguities and errors were
encountered in the during the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the micro-controller state.

@@

% [Trabalhos futuros ]
Future works comprise the development of software with the B method
from functional specification to assembly level, using the Z80 model
presented in this paper. The mechanic compilation from B algorithmic 
constructs to assembly platform is also envisioned.

%  The next steps of the work comprises the development of a real study case in the oil area and the project of a
% compiler to aid the automatization of verification at the assembly level

@@

\paragraph{Acknowledges:}
This work received support from ANP (\textit{Agência Nacional do
  Petróleo, Gás Natural e Biocombustíveis}) and CNPq.
